---
title: "Cover_Data_Analysis"
output: html_document
date: "2024-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Setup

-   Set working directory - must change to location of data on users computer!
-   Load packagesneeded for future code
-   Clear the environment
-   Read in cover data - must change to location of data on users computer!
-   Format date column

```{r setup}
#setting working directory
setwd("C:/Users/edeegan/OneDrive - DOI/Fire_project/Fire_project/")
#loading libraries
library(tidyverse)
library(ggrepel)
library(ordinal)
library(pals)
library(egg)

#clearing environment - fresh start!
rm(list = ls())
#reading in cover data
cover=read.csv("SAGU_data/PSME/PSME_Cover - Species Composition (metric)_XPT.csv", na.strings=c("","NA"))

#formatting date column
cover$Date=as.Date(cover$Date, format="%m/%d/%Y")
#creating new column for just year
cover$Year=str_split_i(cover$Date, "-", 1)
```

## Most Common Herbaceous Species Data

### Setup

-   Filter for most common herbaceous species data
-   Count species observations for each year
-   Calculate percentages
-   Rename species codes to common names
-   must change for local species!
-   Classify as grass or forb
-   Calculating percentages of grasses and forbs

```{r MOST COMMON HERBACEOUS SPECIES DATA}

#uv1 describes category, filtering for most common herbaceous species
common_species=cover[which(cover$UV1=="Most Common Herbaceous Species"),]

#counting species for each year and creating new data frame for results
species_summary=common_species %>% group_by(Year) %>% count(Species.Symbol)
species_summary=species_summary[-which(species_summary$Species.Symbol=="XXXX"),]

#counting number of times a species was recorded for each year to calculate percentages
species_summary <- species_summary %>%
  group_by(Year) %>%
  mutate(total = sum(n)) %>%
  ungroup()

#calculating species percentage of total for each year
species_summary <- species_summary %>% group_by(Species.Symbol) %>%
  mutate(percent = round(n/total*100, 1))


#renaming species code to common name 
species_summary=species_summary %>%
  mutate(Species.Symbol = case_match(Species.Symbol, 
                                 'ARLA1' ~ 'Sandwort', 
                                 'BRCI1' ~ 'Fringed Brome',
                                 'CAGE1' ~ 'White Mountain Sedge',
                                 'CHEI1' ~ 'Lipfern',
                                 'COCO1' ~ 'Coulters Horseweed',
                                 'KOMA1' ~ 'Mountain Junegrass',
                                 'OXAL1' ~ 'Wood Sorrel',
                                 'PTAQ1' ~ 'Bracken Fern',
                                 'SCPA1' ~ 'Figwort',
                                 'COCA1' ~ 'Canadian Horseweed',
                                 'COUM1' ~ 'Bastard Toadflax',
                                 'GNAP1' ~ 'Cudweed',
                                 'HODU1' ~ 'Mountain Spray',
                                 'PIFI1' ~ 'Pinyon Rice Grass',
                                 'AGHE1' ~ 'Fragrant Snakeroot',
                                 'CHGR1' ~ 'Fetid Goosefoot',
                                 'GALI1' ~ 'Common Bedstraw',
                                 'GRAS1' ~ 'Unknown Grass',
                                 'POFE1' ~ 'Fendler Bluegrass',
                                 'SYOR1' ~ 'Mountain Snowberry',
                                 'RONE1' ~ 'New Mexican Locust',
                                 'FORB1' ~ 'Unknown Forb'))


#classifying as grass or forb
species_summary=species_summary %>%
  mutate(Veg_type = case_match(Species.Symbol, 
                                 'Sandwort' ~ 'Forb', 
                                 'Fringed Brome' ~ 'Grass',
                                 'White Mountain Sedge' ~ 'Grass',
                                 'Lipfern' ~ 'Forb',
                                 'Coulters Horseweed' ~ 'Forb',
                                 'Mountain Junegrass' ~ 'Grass',
                                 'Wood Sorrel' ~'Forb',
                                 'Bracken Fern' ~ 'Forb',
                                 'Figwort'~ 'Forb',
                                 'Canadian Horseweed'~ 'Forb',
                                 'Bastard Toadflax'~ 'Forb',
                                 'Cudweed'~ 'Forb',
                                 'Mountain Spray'~ 'Forb',
                                 'Pinyon Rice Grass' ~ 'Grass',
                                 'Fragrant Snakeroot'~ 'Forb',
                                 'Fetid Goosefoot'~ 'Forb',
                                 'Common Bedstraw'~ 'Forb',
                                 'Unknown Grass' ~ 'Grass',
                                 'Fendler Bluegrass' ~ 'Grass',
                                 'Mountain Snowberry'~ 'Forb',
                                 'New Mexican Locust'~ 'Forb',
                                 'Unknown Forb'~ 'Forb'))
#counting for each year for percentages
grassvforb=species_summary %>% group_by(Year) %>% count(Veg_type)
grassvforb <- grassvforb %>%
  group_by(Year) %>%
  mutate(total = sum(n)) %>%
  ungroup()
#calculating percentage of total for each year
grassvforb <- grassvforb %>% group_by(Veg_type) %>%
  mutate(percent = round(n/total*100, 1))

```

### Plot

```{r Most Common Herbaceous Species plot}

#plotting most common herbaceous species for each year in a pie chart with a different color for each species

ggplot(species_summary, aes(x="", y=percent, fill=Species.Symbol)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)+
  facet_wrap(~Year)+
  geom_text(aes(label = paste(percent, "%")), size=2, 
                  position = position_stack(vjust=0.5)) +
  labs(x = NULL, y = NULL, fill = NULL)+theme_bw()+scale_fill_manual(values=as.vector(alphabet(22)))+
  labs(title = "PSME plots most common herbaceous species")+ #change this to reflect monitoring type
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank())

#saving plot in PSME plot folder - change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/most_common_species.png", width=10, height=4)

```

## Canopy Cover Data

### Setup

-   Filter for most canopy cover class data
-   Calculate average canopy cover for each year
-   Run Anova to see difference between years

```{r Canopy Cover Data}
#uv1 describes category, filtering for Canopy Cover Class data
canopy=cover[which(cover$UV1=="Canopy Cover Class"),]

#calculating average canopy cover for each year and creating new data frame with results
canopy_summary=canopy %>% group_by(Year) %>% summarize(average=mean(Cover))

#running an anova to see if there is a significant difference between years for canopy cover
anova(lm(Cover ~ factor(Year), data = canopy))
```

### Plots

```{r Canopy Cover plots}

#plotting average canopy cover in bar graph - doesn't show distribution well
canopy_summary %>% ggplot(aes(Year, average))+geom_col()+ylab("Average Canopy Cover")+theme_classic()

#saving average canopy cover plot in PSME folder - change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/canopy_cover_summary.png")

#plotting canopy cover distribution in a jitter and boxplot - shows distribution better
canopy %>% ggplot(aes(Year, Cover, fill=Year))+geom_boxplot()+geom_jitter()+theme_classic()+ylab("Canopy Cover")

#saving canopy cover boxplot in PSME folder - change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/canopy_cover_jitter.png")

#ploting canopy cover count per plot and year, with a different color for each plot
canopy %>% ggplot(aes(Cover, fill=MacroPlot.Name))+geom_histogram()+
  facet_wrap(~Year)+theme_classic()+
  scale_fill_manual(values=as.vector(ocean.haline(10)))

#saving canopy cover by plot graph - change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/canopy_cover_byplot.png")
```

## Comparing Most Common Herbaceous Species Data with Canopy Cover Data

### Setup

-   REQUIRED - run Most Common Herbaceous Species data setup and Canopy Cover data setup before running this chunk
-   Filter out 2004 for canopy cover

```{r Comparing Most Common Herbaceous Species Data with Canopy Cover Data}
#cutting out 2004 to compare - no data exists in 2004 for most common herbaceous species in PSME plots
canopy_2=canopy[-which(canopy$Year=='2004'),]

```

### Plot

```{r Comparing Most Common Herbaceous Species Data with Canopy Cover Data plot}
#plotting canopy boxplot plot and saving as canopy_graph
canopy_graph=canopy_2 %>% ggplot(aes(Year, Cover, fill=Year))+geom_boxplot()+geom_jitter()+theme_classic()+ylab("Canopy Cover")


# stacked bar plot of veg type forb vs grass count by year
stackedbarplot=ggplot(species_summary, aes(fill=Veg_type, x=Year)) + 
  geom_bar(position="stack", stat="count")+theme_classic()


#stacking stacked bar plot over canopy cover boxplot to compare trends
canopy_species_comparison=ggarrange(stackedbarplot, canopy_graph)
#saving comparison of veg type count and canopy cover- change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/canopy_species_comparison.png", plot=canopy_species_comparison)


# stacked bar plot of veg type forb vs grass percent by year
stackedbarplot_percent=ggplot(grassvforb, aes(fill=Veg_type, x=Year, y=percent)) + 
  geom_bar(position="stack", stat="identity")+theme_classic()
stackedbarplot_percent

#saving stacked bar plot with percentages - change to desired location on computer
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/percent_grassvforb.png")

#stacking stacked bar plot with percentages over canopy cover boxplot to compare trends
ggarrange(stackedbarplot_percent, canopy_graph)
#save if you want to I didn't lol
```

## Additional Species

### Setup

-   Read in cover data again for a fresh start after canopy cover analysis
-   Format cover date column
-   Filter for Additional Species data
-   Remove XXXX species observations
-   Count additional species observations by year and plot

```{r Additional Species}
#rereading in cover data
cover=read.csv("SAGU_data/PSME/PSME_Cover - Species Composition (metric)_XPT.csv", na.strings=c("","NA"))
#formatting date column
cover$Date=as.Date(cover$Date, format="%m/%d/%Y")
#creating new column for just year
cover$Year=str_split_i(cover$Date, "-", 1)

#uv1 describes category, filtering for additional species
additional_species=cover[which(cover$UV1=="Additional Species"),]
#removing XXXX species
additional_species=additional_species[-which(additional_species$Species.Symbol=="XXXX"),]
#counting observations by year and plot
additional_species=additional_species %>% group_by(Year, MacroPlot.Name) %>% count()
```

### Plot

```{r Additional Species plot}
#plotting number of additional species recorded each year and coloring by plot
ggplot(additional_species, aes(x=Year, y=n, fill=MacroPlot.Name))+
  geom_histogram(stat="identity")+ylab("Number of Additional Species recorded")
#saving plot of additional species recorded
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/additionalspecies.png", width=6, height=4)


```

## Invasive Species

### Setup

-   Filter for Invasive Species data
-   Remove XXXX species observations
-   Count Invasive species observations by year and plot

```{r Invasive Species}
#uv1 describes category
invasive_species=cover[which(cover$UV1=="Invasive Species"),]
#removing XXXX species
invasive_species=invasive_species[-which(invasive_species$Species.Symbol=="XXXX"),]
#counting observations by year and plot
invasive_species=invasive_species %>% group_by(Year, MacroPlot.Name) %>% count()
```

### Plot

```{r Invasive Species plot}
#plotting number of invasive species recorded each year and coloring by plot
ggplot(invasive_species, aes(x=Year, y=n, fill=MacroPlot.Name))+
  geom_histogram(stat="identity")+ylab("Number of invasive Species recorded")
#saving plot of invasive species recorded
ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/invasivespecies.png", width=6, height=4)

```
