---
title: "DBH_Analysis"
output: html_document
date: "2024-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Setup

-   Set working directory - must change to location of data on users computer!
-   Load packages needed for future code
-   Clear the environment
-   Read in tree data - must change to location of data on users computer!
-   Format date column
-   Filter for live trees
-   Combine 1990, 1991, and 1992 - specific to Saguaro!

```{r}
#set working directory
setwd("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/")
#load packages
library(tidyverse)
library(ggforce)
library(stringr)
library(dplyr)
#clear the environment - fresh start!
rm(list = ls())

#read in tree data
tree=read.csv("C:/Users/edeegan/OneDrive - DOI/Fire_project/Fire_project/SAGU_data/PSME/PSME_Trees - Individuals (metric)_XPT.csv")
#format date column
tree$Date_format=as.Date(tree$Date, format="%m/%d/%Y")
tree$Year=str_split_i(tree$Date_format, "-", 1)
#filter for live trees
tree=tree[which(tree$Status=="L"),]

#combine 1990, 1991, and 1992
early_ss=tree[which(tree$Year %in% c("1990", "1991", "1992")), ]
tree=tree[-which(tree$Year %in% c("1990", "1991", "1992")), ]
early_ss$Year="1992"
tree=rbind(early_ss,tree)


```

#### Label trees by size class

-   Pole is less than 15.1 dbh
-   Medium is between 15.1 and 30 dbh
-   Overstory is over 30 dbh

```{r label trees}

for(x in 1:nrow(tree)){
  if(is.na(tree[x, "DBH"])){
    tree[x, "SizeClass"]="NA"
  }else{
    if(tree[x, "DBH"]<=15.1){
      #pole tree
      tree[x, "SizeClass"]="Pole (<15.1)"
    }else if(15.1<tree[x, "DBH"] & tree[x, "DBH"]<30){
      #medium tree
      tree[x, "SizeClass"]="Medium (<30)"
    }else{
      #overstory tree
      tree[x, "SizeClass"]="Overstory (30<)"
    }
    
  }
}

```

-   drop blank DBHs
-   round DBHs to whole numbers
-   calculate density in count per acre

```{r}
tree=tree %>% drop_na(DBH)
tree=tree %>% mutate(DBH=round(DBH, 0))

#density calculation
tree_1=tree %>% mutate(density=1/(tree$SubFrac*0.24710538))
```

### Plotting dbh total density

\*warning - this graph is slightly misleading because poles are recorded at a different scale than medium and overstory trees (in Saguaro at least). However, they are all converted to the same unit (stems/acre) so they can be plotted together accurately.

```{r}
ggplot(tree_1, aes(x=DBH, y=density, fill=SizeClass))+
  geom_histogram(stat="identity")+
  facet_grid(~Year)+theme_classic()+ylab("Total Density (stems/acre)")+xlab("DBH")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#saving plot
ggsave("PSME_Plots/dbh_total_density.png", height=4, width=6)
```

### Plot density by size class

\*see above warning

```{r}
#group by size class - count then do density

tree_2=tree %>% mutate(acres=0.24710538*tree$SubFrac) %>% group_by(Year, SizeClass, acres, MacroPlot.Name) %>% count() %>% mutate(density=n/acres)
tree_2=tree_2 %>% group_by(Year, SizeClass) %>% summarize_at(vars(density), list(average_density=mean))
level_order <- c('Pole (<15.1)', 'Medium (<30)', 'Overstory (30<)') 

ggplot(tree_2, aes(x=factor(SizeClass, level_order), y=average_density, fill=SizeClass))+
  geom_histogram(stat="identity")+
  facet_grid(~Year)+theme_classic()+ylab("Average Density (stems/acre)")+xlab("Size Class")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggsave("PSME_Finalized_Plots/Dead_tree_size_histogram.png", width=15, height=5)

```

### Plot density by MacroPlot

\*see above warning

```{r}

#group by plot - count then do density

tree_3=tree %>% group_by(Year, SizeClass, MacroPlot.Name, SubFrac) %>% count() %>% mutate(acres=0.24710538*SubFrac) %>% mutate(density=n/acres)
level_order <- c('Pole (<15.1)', 'Medium (<30)', 'Overstory (30<)') 

ggplot(tree_3, aes(x=factor(MacroPlot.Name), y=density, fill=SizeClass))+
  geom_histogram(stat="identity")+
  facet_grid(~Year)+theme_classic()+ylab("Density (stems/acre)")+xlab("Size Class")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


ggsave("PSME_Finalized_Plots/Dead_tree_size_histogram.png", width=15, height=5)
```

### Additional Density Graphs

-   density sum per year
-   box plots separated by size class

```{r}

tree_4=tree %>% group_by(Year, SizeClass, MacroPlot.Name, SubFrac) %>% count() %>% 
  mutate(acres=0.24710538*SubFrac) %>% mutate(density=n/acres) %>%
  group_by(Year, SizeClass) %>% summarize(density_sum=sum(density))


tree_4 %>% ggplot(aes(x=Year, y=density_sum))+geom_bar(stat="identity")

tree_5=tree %>% group_by(Year, SizeClass, MacroPlot.Name, SubFrac) %>% count() %>% 
  mutate(acres=0.24710538*SubFrac) %>% mutate(density=n/acres)

tree_5 %>% ggplot(aes(x=Year, y=density, fill=factor(SizeClass, level=level_order)))+geom_boxplot()
                
  
```
