---
title: "Tree_Basal_Area_Analysis"
output: html_document
date: "2024-05-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Setup

-   Set working directory - must change to location of data on users computer!
-   Load packages needed for future code
-   Clear the environment
-   Read in cover data - must change to location of data on users computer! - specific to Saguaro\*
-   Filter for canopy cover data - specific to Saguaro\*
-   Format date column
-   Read in tree data - must change to location of data on users computer!
-   Format date column
-   Combining years 1990, 1991, 1992 - specific to Saguaro
-   Renaming species code to common name for tree species - must change for local species!

\*Note for the cover data - this is a Saguaro specific protocol and won't apply to data from other parks. It is included in this script because we compared our canopy cover data to tree basal area. The protocol has since been discontinued. Skip these sections if you are not from Saguaro

```{r setup}
#set working directory  
setwd("C:/Users/edeegan/OneDrive - DOI/Fire_project/Fire_project/")
#load packages
library(tidyverse)
library(ggrepel)
library(multcompView)
library(ggridges)
library(viridis)
#clearing environment - fresh start!
rm(list = ls())

#reading in cover data
cover=read.csv("SAGU_data/PSME/PSME_Cover - Species Composition (metric)_XPT.csv", na.strings=c("","NA"))
#formatting date column
cover$Date=as.Date(cover$Date, format="%m/%d/%Y")
#creating new column for just year
cover$Year=str_split_i(cover$Date, "-", 1)
#uv1 describes category
canopy=cover[which(cover$UV1=="Canopy Cover Class"),]

#reading in tree data
tree=read.csv("C:/Users/edeegan/OneDrive - DOI/Fire_project/Fire_project/SAGU_data/PSME/PSME_Trees - Individuals (metric)_XPT.csv")
#formatting date column
tree$Date=as.Date(tree$Date, format="%m/%d/%Y")
#creating new column for just year
tree$Year=str_split_i(tree$Date, "-", 1)

#combining 1990, 1991, 1992
tree[which(tree$Year=='1990' | tree$Year=='1991'), "Year"]='1992'

#renaming species code to common name - CHANGE FOR LOCAL SPECIES
tree=tree %>%
  mutate(Species.Symbol = case_match(Species.Symbol, 
                                 'ABCO1' ~ 'White Fir', 
                                 'QUGA1' ~ 'Gambels Oak',
                                 'PIPO1' ~ 'Ponderosa Pine',
                                 'PIST1' ~ 'Southwestern White Pine',
                                 'PSME1' ~ 'Douglas Fir'))


```

## Basal Area Calculation

### Setup

-   only live trees
-   option to remove poles - un-comment the second line

```{r basal area}

#TREE BASAL AREA CALCULATION
tree_1=tree[-which(tree$Status=="D"),] #only live trees
#tree_1=tree_1[-which(tree_1$DBH<=15.1),] #remove poles - uncomment to implement!

tree_1$DBH=tree_1$DBH*0.39370079 #convert to inches
tree_1$basal_area=(tree_1$DBH^2)*0.005454 #formula
tree_1$basal_area=as.numeric(tree_1$basal_area)
tree_1=tree_1[-which(is.na(tree_1$basal_area)),] #only tree w basal area



```

## Plotting Basal Area

### Plot Basal Area by Species

```{r plot}

#sum basal area by year and plot
tree_basal=tree_1 %>% group_by(Year, MacroPlot.Name, SubFrac, Species.Symbol) %>% summarize(total_ba=sum(basal_area))
#adding acres
tree_basal=tree_basal %>%
  mutate(acres = 0.24710538*SubFrac) %>% mutate(density=total_ba/acres)
#sum density by plot and year
tree_basal=tree_basal %>% group_by(Year, MacroPlot.Name, Species.Symbol) %>% summarize(totalba_peracre=sum(density))

#averaging/plot by year
tree_basal_summary=tree_basal %>% group_by(Year, Species.Symbol) %>% summarize(avg_plot_ba=mean(totalba_peracre))

#plotting by species
plot=ggplot(tree_basal, aes(x=Year, y=totalba_peracre, fill=Species.Symbol))+
  geom_bar(position="stack", stat="identity")+theme_classic()+ylab("Total Basal Area per Acre (ft^2/acre)")

#Change for year of fire!!
plot=plot +annotate("text", x="2003", y=50,size=3, label="Fire", color="red")
plot

ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/tree_basal_area.png", width=7, height=3)


```

### Plot Basal Area by plot

```{r plot}
#sum basal area by year and plot
tree_basal=tree_1 %>% group_by(Year, MacroPlot.Name, SubFrac) %>% summarize(total_ba=sum(basal_area))
#adding acres
tree_basal=tree_basal %>%
  mutate(acres = 0.24710538*SubFrac) %>% mutate(density=total_ba/acres)
#sum density by plot and year
tree_basal=tree_basal %>% group_by(Year, MacroPlot.Name) %>% summarize(totalba_peracre=sum(density))

#averaging/plot by year
tree_basal_summary=tree_basal %>% group_by(Year) %>% summarize(avg_plot_ba=mean(totalba_peracre))

#plot
ggplot(tree_basal, aes(x=totalba_peracre, fill=MacroPlot.Name))+geom_histogram()+facet_grid(rows=vars(Year))+theme_classic()

ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/basalarea_byplot.png", width=7, height=4)

```

Create labels for years that are significantly different from each other

```{r}

basal_a_model=aov(totalba_peracre ~ as.factor(Year), data=tree_basal)
anova(basal_a_model)
tukey=TukeyHSD(basal_a_model, conf.level=.95)


# I need to group the treatments that are not different each other together.
generate_label_df <- function(TUKEY, variable){
  
  # Extract labels and factor levels from Tukey post-hoc 
  Tukey.levels <- TUKEY[[variable]][,4]
  Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])
  
  #I need to put the labels in the same order as in the boxplot :
  Tukey.labels$treatment=rownames(Tukey.labels)
  Tukey.labels=Tukey.labels[order(Tukey.labels$treatment) , ]
  return(Tukey.labels)
}

#run function
labels=generate_label_df(tukey, "as.factor(Year)")
#print labels
print(labels)

```

### Density ridge graph

```{r plot}
ggplot(tree_basal, aes(x=totalba_peracre, y=Year, fill=after_stat(x)))+geom_density_ridges_gradient()+
  scale_colour_gradient(low = "#132B43",
                        high = "#56B1F7")+theme_classic()+
  scale_y_discrete(limits=rev)

ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/basalarea_distribution.png", width=7, height=4)


```

### Plotting basal area against canopy cover data

```{r}

#plotting basal area against canopy cover data
#need average canopy cover per plot

canopy=canopy %>% group_by(Year, MacroPlot.Name) %>% summarize(cover=mean(Cover))

canopy_basal=inner_join(canopy, tree_basal, by = join_by(Year, MacroPlot.Name))

ggplot(canopy_basal, aes(x=cover, y=totalba_peracre))+
  geom_point()+geom_smooth(method="lm")+theme_classic()

ggsave("C:/Users/edeegan/OneDrive - DOI/FFIanalysis/Analysis/PSME_Plots/basalarea_v_cover.png", width=5, height=4)

model <- lm(totalba_peracre~cover, data=canopy_basal)
summary(model)
```
